---
title: "Predicting readmission probability for diabetes inpatients"
date: ' '
output:
  pdf_document:
    toc: yes
    toc_depth: '4'
  html_document:
    code_folding: show
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '4'
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width = 7, fig.height = 4)
if(!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr, ggplot2, glmnet, car, data.table, ggplot, plotly)   #add your packages here
```

#Loading data and recoding the last column
```{r}
data.fl <- read.csv("readmission.csv",header=T, na.string = c("","?"))
data <- data.fl
names(data)
```
#Cheking for missing values
```{r}
sum(is.na(data))
```
```{r}
data <- na.omit(data)
sum(is.na(data))
dim(data)
```
#Analyzing the data types
```{r}
str(data)
```
#Cheking readmitted col
```{r}
unique(data$readmitted)
head(data$readmitted)
```
#Recoding this col
```{r}
data$readmitted <- ifelse((data$readmitted == ">30") | (data$readmitted == "<30"), "YES", data$readmitted)
head(data$readmitted)
```
#Checking whether chr data types are truly factors or not
```{r}
unique(data$diag2_mod)
```
#Above analysis suggests that (as an example) diag2_mod has 25 unique values out of 100,000+ observations, so we might want to treat it as a factor. This applies to other chr data types also

#Convering chr to factors
```{r}
data$readmitted <- as.factor(data$readmitted)
#data$readmitted <- (ifelse(data$readmitted =="YES", "1", "0"))
str(data$readmitted)
```
#Splitting the data
```{r}
data.fl <- data

numl <- 0.02 #using this low number because it otherwise is becoming very computationally intensive

#N <- length(data$readmitted)
#n1 <- floor(numl*N)
#set.seed(10)

#idx_train <- sample(N, n1)
#idx_no_train <- (which(! seq(1:N) %in% idx_train))
#idx_val <- which(! idx_train %in% idx_no_train)
#data <- data.fl[idx_train,]
#data.val <- data.fl[idx_val,]


# Split the data:
N <- length(data$readmitted)
n1 <- floor(.6*N)
n2 <- floor(.2*N)
set.seed(10)

# Split data to three portions of .6, .2 and .2 of data size N
idx_train <- sample(N, n1)
idx_no_train <- (which(! seq(1:N) %in% idx_train))
idx_test <- sample( idx_no_train, n2)
idx_val <- which(! idx_no_train %in% idx_test)
data.train <- data.fl[idx_train,]
data.test <- data.fl[idx_test,]
data.val <- data.fl[idx_val,]

data <-rbind(data.train,data.test)

```

1) Identify important factors that capture the chance of a readmission within 30 days.

**Step I: Prepare the design matrix and response**
```{r}
Y <- data$readmitted
X <- model.matrix(readmitted~., data=data)[,-1]
dim(X)
```
**Step II: Select a lasso model** 
```{r results=TRUE}
set.seed(10)
fit1.cv <- cv.glmnet(X, Y, alpha=1, family="binomial", nfolds = 10, type.measure = "deviance")  
plot(fit1.cv)
```
#Model based on lambda.1se
```{r results=TRUE}
coef.1se <- coef(fit1.cv, s="lambda.1se")  
coef.1se <- coef.1se[which(coef.1se !=0),] 
as.matrix(coef.1se)
```
#Model based on lambda.min
```{r results=TRUE}
coef.min <- coef(fit1.cv, s="lambda.min") 
coef.min <- coef.min[which(coef.min !=0), ]
as.matrix(coef.min)
```
# Both the models use all the variables, and appear to be the same.

2) For the purpose of classification, propose a model that can be used to predict whether a patient will be a readmit within 30 days. Justify your choice. Hint: use a decision criterion, such as AUC, to choose among a few candidate models.

#Fitting model using beta.1se
```{r}
fit.logit.1 <- glm(readmitted~ encounter_id+patient_nbr+race+gender+time_in_hospital+num_procedures+num_medications+number_outpatient+number_emergency+number_inpatient+number_diagnoses+max_glu_serum+A1Cresult+metformin+glimepiride+glipizide+glyburide+pioglitazone+rosiglitazone+insulin+change+diabetesMed+disch_disp_modified+adm_src_mod+adm_typ_mod+age_mod+diag1_mod+diag2_mod+diag3_mod, family=binomial, data=data)
Anova(fit.logit.1)
```

#Fitting model using beta.min (This and prior model differ in just 1 variable)
```{r}
fit.logit.2 <- glm(readmitted~ encounter_id+patient_nbr+race+gender+time_in_hospital+num_lab_procedures+num_procedures+num_medications+number_outpatient+number_emergency+number_inpatient+number_diagnoses+max_glu_serum+A1Cresult+metformin+glimepiride+glipizide+glyburide+pioglitazone+rosiglitazone+insulin+change+diabetesMed+disch_disp_modified+adm_src_mod+adm_typ_mod+age_mod+diag1_mod+diag2_mod+diag3_mod, family=binomial, data=data)
Anova(fit.logit.2)
```

#Using AUC to compare the two models
```{r}
fit.logit.1.1 <- glm(readmitted~ encounter_id+patient_nbr+race+gender+time_in_hospital+num_procedures+num_medications+number_outpatient+number_emergency+number_inpatient+number_diagnoses+max_glu_serum+A1Cresult+metformin+glimepiride+glipizide+glyburide+pioglitazone+rosiglitazone+insulin+change+diabetesMed+disch_disp_modified+adm_src_mod+adm_typ_mod+age_mod+diag1_mod+diag2_mod+diag3_mod, family=binomial(logit), data=data)

fit.logit.2.1 <- glm(readmitted~ encounter_id+patient_nbr+race+gender+time_in_hospital+num_lab_procedures+num_procedures+num_medications+number_outpatient+number_emergency+number_inpatient+number_diagnoses+max_glu_serum+A1Cresult+metformin+glimepiride+glipizide+glyburide+pioglitazone+rosiglitazone+insulin+change+diabetesMed+disch_disp_modified+adm_src_mod+adm_typ_mod+age_mod+diag1_mod+diag2_mod+diag3_mod, family=binomial(logit), data=data)

fit.logit.1.roc <- roc(data$readmitted, fit.logit.1$fitted)
fit.logit.2.roc <- roc(data$readmitted, fit.logit.2$fitted)
fit.logit.1.1.roc <- roc(data$readmitted, fit.logit.1.1$fitted)
fit.logit.2.1.roc <- roc(data$readmitted, fit.logit.2.1$fitted)
```

#Plotting them
```{r}
plot(1-fit.logit.1.roc$specificities,
fit.logit.1.roc$sensitivities, col="red", lwd=3, type="l",
xlab="False Positive",
ylab="Sensitivity")
lines(1-fit.logit.2.roc$specificities, fit.logit.2.roc$sensitivities, col="blue", lwd=3)
lines(1-fit.logit.1.1.roc$specificities, fit.logit.1.1.roc$sensitivities, col="green", lwd=3)
lines(1-fit.logit.2.1.roc$specificities, fit.logit.2.1.roc$sensitivities, col="orange", lwd=3)
legend("bottomright",
c(paste0("fit.logit.1 AUC=", round(fit.logit.1.roc$auc,2)),
paste0("fit.logit.2 AUC=", round(fit.logit.2.roc$auc, 2)),
paste0("fit.logit.1.1 AUC=", round(fit.logit.1.1.roc$auc, 2)),
paste0("fit.logit.2.1 AUC=", round(fit.logit.2.1.roc$auc, 2))),
col=c("red", "blue", "green","orange"),
lty=1)
```

#These models are nearly the same, so we work with the simplest possible model below.
```{r}
Anova(fit.logit.1.1)
```
#Fitting and comparing a reduced model, removed "change"
```{r}
fit.logit.1.1.2 <- glm(readmitted~ encounter_id+patient_nbr+race+gender+time_in_hospital+num_procedures+num_medications+number_outpatient+number_emergency+number_inpatient+number_diagnoses+max_glu_serum+A1Cresult+metformin+glimepiride+glipizide+glyburide+pioglitazone+rosiglitazone+insulin+diabetesMed+disch_disp_modified+adm_src_mod+adm_typ_mod+age_mod+diag1_mod+diag2_mod+diag3_mod, family=binomial(logit), data=data)

anova(fit.logit.1.1.2, fit.logit.1.1, test="Chisq")
```
```{r}
Anova(fit.logit.1.1.2)
```
#Fitting and comparing a reduced model, removed "pioglitazone" and "max_glu_serum"
```{r}
fit.logit.1.1.3 <- glm(readmitted~ encounter_id+patient_nbr+race+gender+time_in_hospital+num_procedures+num_medications+number_outpatient+number_emergency+number_inpatient+number_diagnoses+A1Cresult+metformin+glimepiride+glipizide+glyburide+rosiglitazone+insulin+diabetesMed+disch_disp_modified+adm_src_mod+adm_typ_mod+age_mod+diag1_mod+diag2_mod+diag3_mod, family=binomial(logit), data=data)

anova(fit.logit.1.1.3, fit.logit.1.1.2, test="Chisq")
```
```{r}
Anova(fit.logit.1.1.3)
```
#Fitting and comparing a reduced model, removed "glyburide"
```{r}
fit.logit.1.1.4 <- glm(readmitted~ encounter_id+patient_nbr+race+gender+time_in_hospital+num_procedures+num_medications+number_outpatient+number_emergency+number_inpatient+number_diagnoses+A1Cresult+metformin+glimepiride+glipizide+rosiglitazone+insulin+diabetesMed+disch_disp_modified+adm_src_mod+adm_typ_mod+age_mod+diag1_mod+diag2_mod+diag3_mod, family=binomial(logit), data=data)

anova(fit.logit.1.1.4, fit.logit.1.1.3, test="Chisq")
```
```{r}
Anova(fit.logit.1.1.4)
```
#Fitting and comparing a reduced model i.e., removing "glimepiride" led to a reduction in model quality as measured by p-value, so we stop here and take the previous fit.logit.1.1.4 as our final model.

3) Based on a quick and somewhat arbitrary guess, we estimate **it costs twice as much** to mislabel a readmission than it does to mislabel a non-readmission. Based on this risk ratio, propose a specific classification rule to minimize the cost. 

#P(Y=Yes|x)>0.5/(1+0.5)=1/3
#logit > log((1/3)/(2/3))= -0.6931


4) We suggest you to split the data first to Training/Testing/Validation data:

- Use training/testing data to land a final model (If you only use LASSO to land a final model, we will not need testing data since all the decisions are made with cross-validations.)

#Yes, we only used LASSO. So to begin with we combined the training and the testing data to build the model and use cross validation.

- Evaluate the final model with the validation data to give an honest assessment of your final model.
```{r}
predictions <- predict(fit.logit.1.1.4, newdata = data.val, type = "response")
predicted.classes <- as.factor(ifelse(predictions > 0.334, "2", "1"))
MCE.bayes <- (5*sum(predictions[data.val$readmitted == "2"] != "2")
              + sum(predictions[data.val$readmitted == "1"] != "1"))/length(data.val$readmitted)
MCE.bayes
```

